---
title: "EA_cleaning"
format: html
author: Savannah Kjaer
date: 04-15-2026
---

## Set up

### load packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(readxl)
library(here)
library(janitor)
```

## Aboveground

### read in data

```{r, message=FALSE}
cn <- read_excel("~/LTER-pluck/2024_pluck/raw_data/CN_results2.xlsx")
view(cn)
#1007 of 17 variables

id <- read_csv("~/LTER-pluck/2024_pluck/processed_data/ag_pluck_id.csv")
#1134 of 7
```

### tidy columns and rows

```{r, message=FALSE}
cn <- clean_names(cn)

colnames(cn)
# removing unnecessary columns
cn1 <- select(cn, -c("group", "sample_name", "inj_date", "inj_time", "type", "weight_mg", "protein_factor", "humidity_percent", "as", "vial"))
#of 6

cn1$carbon <- as.numeric(cn1$carbon)
cn1$nitrogen <- as.numeric(cn1$nitrogen)
cn1$filename <- as.character(cn1$filename)

names(cn1)[names(cn1) == "id"] <- "plot_tissue_id"
```

### separate AG masses from others

```{r}
cn_ag <- filter(cn1, sample_type == "AG")
view(cn_ag)
#760
```

### corrections

```{r}
# two ids from AG mass data so one renamed to another quad id
cn_ag$plot_tissue_id[cn_ag$filename == "09042025013"] <- "1601"

# looking into ids that appeared in EA results that are not id names
# 602, 685, 1548, 1549, 1084, 1355
# 602 was mistyped, paper datasheet say 662
# 685 was mistyped, paper datasheet say 658
# 1548 was mistyped, paper datasheet say 1506
# 1549 was mistyped, paper datasheet say 1507
# 223 was mistyped, paper datasheet say 222
# 1084 unclear what sample it was
# 1355 is under the id 1164, which was missing so showed up as an id but not a sample

cn_ag$plot_tissue_id[cn_ag$filename == "04042025044"] <- "662"
cn_ag$plot_tissue_id[cn_ag$filename == "04042025042"] <- "658"
cn_ag$plot_tissue_id[cn_ag$filename == "09042025017"] <- "1506"
cn_ag$plot_tissue_id[cn_ag$filename == "09042025018"] <- "1507"
cn_ag$plot_tissue_id[cn_ag$filename == "03042025105"] <- "222"
cn_ag$plot_tissue_id[cn_ag$filename == "24032025073"] <- "NA"
cn_ag$plot_tissue_id[cn_ag$filename == "23012026046"] <- "1164"
```

### combine

```{r, message=FALSE}
head(cn_ag)
#760 obs of 7

head(id)
#1134 of 7 
id$plot_tissue_id <- as.character(id$plot_tissue_id)

cn2 <- left_join(cn_ag, id)
# Joining with `by = join_by(plot_tissue_id)`

view(cn2)
#760 of 13
```

### error check

```{r, message=FALSE}
cn2.na <- subset(cn2, (is.na(cn2$block)))
#117 obs

# remove check results
cn3 <- cn2 %>%
  filter(!grepl("Check", plot_tissue_id))
#697

# "NA" wasn't NA
cn3$plot_tissue_id[cn3$plot_tissue_id == "NA"] <- NA

cn3.na <- subset(cn3, (is.na(cn3$block)))
#54
# 1186, 538, 1617, 1488, 1476
# these are single quads that weren't included on the first run, ran as a single quad later
# leave out for now, don't want to over-repesent these quads

cn4 <- cn3 %>%
  filter(!(block %in% c(NA)))
#643, good!!
```

### burnt samples

Looked at values of burnt against unburnt samples, not terribly different but I think it's best to remove them entirely, rather than keep potentially flawed samples. This removes all samples from B1 F0.5, and B1 CT Q3&4.

```{r}
# checking burnt B1 against unburnt B2&3
test <- cn4 %>%
  filter(treatment %in% c("F0.5"))

test %>%
  ggplot() +
  geom_boxplot(aes(x=tissue, y=carbon, col=notes)) + 
  facet_wrap(~species)
test %>%
  ggplot() +
  geom_boxplot(aes(x=tissue, y=nitrogen, col=notes)) + 
  facet_wrap(~species)
```

### remove burnt samples

```{r}
burnt <- cn4 %>%
  filter(notes %in% c("burnt sample"))
#40
# all burnt, should be removed

cn5 <- cn4 %>%
  filter(!(notes %in% c("burnt sample")))

view(cn5)
#603 obs
```

### create growth form column

```{r}
#create Growth_form column
cn6 <- cn5 %>% 
  mutate(growth_form = case_when(
    startsWith(species, "Eriophorum vaginatum") ~ "Graminoid",
    startsWith(species, "Other Graminoid") ~ "Graminoid",
    startsWith(species, "Carex bigelowii") ~ "Graminoid",
    startsWith(species, "Betula nana") ~ "Deciduous",
    startsWith(species, "Salix pulchra") ~ "Deciduous",
    startsWith(species, "Vaccinium uliginosum") ~ "Deciduous",
    startsWith(species, "Other Deciduous") ~ "Deciduous",
    startsWith(species, "Ledum palustre") ~ "Evergreen",
    startsWith(species, "Vaccinium vitis-idaea") ~ "Evergreen",
    startsWith(species, "Cassiope tetragona") ~ "Evergreen",
    startsWith(species, "Other Evergreen") ~ "Evergreen",
    startsWith(species, "Rubus chamaemorus") ~ "Forb",
    startsWith(species, "Petasites frigidus") ~ "Forb",
    startsWith(species, "Other Forb") ~ "Forb",
    startsWith(species, "Sphagnum") ~ "Moss",
    startsWith(species, "Other Mosses") ~ "Moss",
    startsWith(species, "Mixed Lichen") ~ "Lichen",
    startsWith(species, "Litter") ~ "Litter"
    ))
#of 14
```

### final tidy

```{r}
head(cn6)

cn7 <- select(cn6, -c("filename", "date", "sample_type", "surface_area_cm2"))
col_order <- c("site", "block", "treatment",  "species", "tissue", "growth_form", "plot_tissue_id", "nitrogen", "carbon",  "notes")
cn7 <- cn7[, col_order]


names(cn7)[names(cn7) == 'nitrogen'] <- "nitrogen_percent"
names(cn7)[names(cn7) == 'carbon'] <- "carbon_percent"
```

### write final_data/ag_cn.csv

```{r}
view(cn7)
#603 of 10

write.csv(cn7, "~/LTER-pluck/2024_pluck/processed_data/final_data/ag_cn.csv", row.names=FALSE)
```

## Belowground

### read in data

```{r, message=FALSE}
id_bg <- read_csv("~/LTER-pluck/2024_pluck/processed_data/bg_pluck_id.csv")
#108 of 6
```

### separate BG masses from others

```{r}
names(cn1)[names(cn1) == "plot_tissue_id"] <- "plot_root_id"

cn_bg <- filter(cn1, sample_type == "BG")
view(cn_bg)
#80
```

### corrections

```{r}
# looking into ids that appeared in EA results that are not id names
# 395 unclear what sample it was

cn_bg$plot_root_id[cn_bg$filename == "11042025025"] <- "NA"
cn_bg$plot_root_id[cn_bg$filename == "24032025041"] <- "175"
```

### combine

```{r, message=FALSE}
head(id_bg)
#108 of 6
id_bg$plot_root_id <- as.character(id_bg$plot_root_id)

cn_bg2 <- left_join(cn_bg, id_bg)
# Joining with `by = join_by(plot_root_id)`

view(cn_bg2)
#80 of 12
```

### error check

```{r, message=FALSE}
cn2.na <- subset(cn_bg2, (is.na(cn_bg2$block)))
#7 obs

# remove check results
cn_bg3 <- cn_bg2 %>%
  filter(!grepl("Check", plot_root_id))
#74

# "NA" wasn't NA
cn_bg3$plot_root_id[cn_bg3$plot_root_id == "NA"] <- NA

cn_bg3.na <- subset(cn_bg3, (is.na(cn_bg3$plot_root_id)))
#1
# checked, should be removed 

cn_bg4 <- cn_bg3 %>%
  filter(!(plot_root_id %in% c(NA)))
#73, good!!

cn_bg4.na <- subset(cn_bg4, (is.na(cn_bg4$block)))
```

### final tidy

```{r}
head(cn_bg4)

cn_bg5 <- select(cn_bg4, -c("filename", "date", "sample_type"))
col_order <- c("site", "block", "treatment", "horizon", "root_type", "plot_root_id", "nitrogen", "carbon", "notes")
cn_bg5 <- cn_bg5[, col_order]

names(cn_bg5)[names(cn_bg5) == 'nitrogen'] <- "nitrogen_percent"
names(cn_bg5)[names(cn_bg5) == 'carbon'] <- "carbon_percent"
```

### write final_data/bg_cn.csv

```{r}
view(cn_bg5)
#73 of 9

write.csv(cn_bg5, "~/LTER-pluck/2024_pluck/processed_data/final_data/bg_cn.csv", row.names=FALSE)
```

## Total Cores

### separate TC masses from others

```{r}
names(cn1)[names(cn1) == "plot_root_id"] <- "plot_horizon_id"

cn_tc <- filter(cn1, sample_type == "Total Cores")
view(cn_tc)
#44
```

### corrections

```{r}
cn_tc$plot_horizon_id <- gsub("-T", "-M", cn_tc$plot_horizon_id)
# not sure why mineral was labeled as T?

# create block, trt, horizon columns from id
cn_tc1 <- cn_tc %>% 
  separate(plot_horizon_id, c("Block", "Treatment", "Horizon"), sep = "-")
```

### error check

```{r, message=FALSE}
# remove check results
cn_tc2 <- cn_tc1 %>%
  filter(!grepl("Check", Block))
#42

# "NA" wasn't NA
cn_tc2$Block[cn_tc2$Block == "NA"] <- NA

cn_tc2.na <- subset(cn_tc2, (is.na(cn_tc2$Block)))
#2
# checked, should be removed 

cn_tc3 <- cn_tc2 %>%
  filter(!(Block %in% c(NA)))
#40, good!!

cn_tc3 <- clean_names(cn_tc3)
```

### final tidy

```{r}
head(cn_tc3)

cn_tc3$site <- c("MAT06")

cn_tc4 <- select(cn_tc3, -c("filename", "date", "sample_type"))

col_order <- c("site", "block", "treatment", "horizon", "nitrogen", "carbon", "notes")
cn_tc4 <- cn_tc4[, col_order]

names(cn_tc4)[names(cn_tc4) == 'nitrogen'] <- "nitrogen_percent"
names(cn_tc4)[names(cn_tc4) == 'carbon'] <- "carbon_percent"

head(cn_tc4)
#of 7
```

### write final_data/tc_cn.csv

```{r}
view(cn_tc4)
#40 of 7

write.csv(cn_tc4, "~/LTER-pluck/2024_pluck/processed_data/final_data/tc_cn.csv", row.names=FALSE)
```
