---
title: "Pluck cleaning"
format: html
author: Savannah Kjaer
date: 03-09-2026
---

## Set up

### load packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(readxl)
library(here)
library(janitor)
```

## Aboveground

### read in data

```{r, message=FALSE, warning=FALSE}
# read in files from /raw_data/ folder
mass <- read_excel("~/LTER-pluck/2024_pluck/raw_data/Shoot_data_collection_mass.xlsx")
#74 obs, 141 variables
head(mass)
```

### tidy columns and rows

```{r, message=FALSE}
# drop columns we don't want, organizational not informational
mass1 <- select(mass, 
                -c("Date_sorted","Plucker","Balance","Weigher","Notes"))
#74 obs, 136 variables

# drop empty rows at the bottom
mass2 <- mass1[-c(73:74),]
#72 obs of 136 variables

# add site column
mass2$site <- c("MAT06")
#of 137
```

### separate mass and IDs

```{r}
# remove mass columns
mass_id <- select(mass2, -ends_with("mass_g"))
#72 obs, 74 variables

# remove other species column
mass_id1 <- select(mass_id, -ends_with("Species"), -c("Date_collected"))
#72 obs, 68 variables

# remove name columns
mass_g <- select(mass2, -ends_with("bag_number"))
#72 obs, 74 variables
mass_g1 <- select(mass_g, -ends_with("Species"))
#69 variables
```

### classes to numeric

```{r, warning=FALSE, output=FALSE}
# some mass columns as characters not numbers
# turns the "missing" values to NA's

mass_g1 <- mass_g1 %>% mutate(across(ends_with("_mass_g"), as.numeric))

# check
sapply(mass_g1, class)
```

### create longform mass table for publishing

```{r}
# wide2long
mass_g2 <- mass_g1 %>%
  pivot_longer(cols = ends_with("_mass_g"),
               names_to = "tissue",
               values_to = "mass_g")
#4536 obs, 8 variables

# remove "_mass_g" from tissue col
mass_g2$tissue <- 
  sub("_mass_g","",mass_g2$tissue)

# separate values in tissue column into species and tissue
mass_g3 <- mass_g2 %>%
  separate(tissue, c("species", "tissue"))
#4536 of 9

unique(mass_g3$species)
unique(mass_g3$tissue)
```

### sort IDs into one per plot

```{r}
# wide2long
mass_id2 <- mass_id1 %>% 
  pivot_longer(cols = c(ends_with('_bag_number')),
               names_to = "tissue",
               values_to = "plot_tissue_ID")
#4536 obs, 7 variables

# put all quads into one row per plot
mass_id3<-mass_id2 %>%
  pivot_wider(names_from = Quad,
              values_from = plot_tissue_ID)
#1134 of 9

# create one ID per plot, ignore quads

# "NA" wasn't NA
mass_id4 <- mass_id3 %>%
  mutate(across(c(`1`, `2`, `3`, `4`), ~na_if(., "NA")))

# ID is the first appearance of a bag number
# if Q1 exists, uses that number, then tries Q2, Q3, Q4
mass_id4$plot_tissue_ID <- coalesce(mass_id4$`1`, mass_id4$`2`, mass_id4$`3`, mass_id4$`4`)

# remove "_bag_number" in tissue column
mass_id4$tissue <- 
  sub("_bag_number","",mass_id4$tissue) 

# separate values in tissue column into species and tissue
mass_id5 <- mass_id4 %>%
  separate(tissue, c("species", "tissue"))
#1134 of 11

# remove quads
mass_id6 <- mass_id5 %>% 
  select(-c('1', '2', '3', '4'))
#of 7

# now just one ID number for a plot
head(mass_id6)
```

### corrections

```{r}
# these ID's only correspond with B1 CT Q3&4, which were the burnt samples that will be excluded
# make them NA's for Q1&2
mass_id6$plot_tissue_ID[mass_id6$plot_tissue_ID == "204"] <- NA
mass_id6$plot_tissue_ID[mass_id6$plot_tissue_ID == "205"] <- NA
mass_id6$plot_tissue_ID[mass_id6$plot_tissue_ID == "258"] <- NA
mass_id6$plot_tissue_ID[mass_id6$plot_tissue_ID == "59"] <- NA

# two IDs from AG mass data so one renamed to another quad ID
mass_id6$plot_tissue_ID[mass_id6$plot_tissue_ID == "1152" & mass_id6$Treatment == "F5"] <- "1601"
```

### clean column names

```{r}
mass_id6 <- clean_names(mass_id6)
mass_g3 <- clean_names(mass_g3)
```

### expand species

```{r}
# from species code to full species name, also matches 2012 data

unique(mass_g3$species)

mass_g3$species <- mass_g3$species %>% 
  recode(
   "Erivag" = "Eriophorum vaginatum",
     "OtherGraminoid" = "Other Graminoid",
     "Carbig" = "Carex bigelowii",
     "Betnan" = "Betula nana",
     "Salpul" = "Salix pulchra",
     "Vaculi" = "Vaccinium uliginosum",
     "OtherDeciduous" = "Other Deciduous",
     "Ledpal" = "Ledum palustre",
     "Vacvit" = "Vaccinium vitis-idaea",
     "Castet" = "Cassiope tetragona",
     "OtherEvergreen" = "Other Evergreen",
     "Rubcha" = "Rubus chamaemorus",
     "Petfri" = "Petasites frigidus",
     "OtherForb" = "Other Forb",
     "OtherMoss" = "Other Mosses",
     "Lichen" = "Mixed Lichen",
     "Attached" = "Litter"
    )

mass_id6$species <- mass_id6$species %>% 
  recode(
   "Erivag" = "Eriophorum vaginatum",
     "OtherGraminoid" = "Other Graminoid",
     "Carbig" = "Carex bigelowii",
     "Betnan" = "Betula nana",
     "Salpul" = "Salix pulchra",
     "Vaculi" = "Vaccinium uliginosum",
     "OtherDeciduous" = "Other Deciduous",
     "Ledpal" = "Ledum palustre",
     "Vacvit" = "Vaccinium vitis-idaea",
     "Castet" = "Cassiope tetragona",
     "OtherEvergreen" = "Other Evergreen",
     "Rubcha" = "Rubus chamaemorus",
     "Petfri" = "Petasites frigidus",
     "OtherForb" = "Other Forb",
     "OtherMoss" = "Other Mosses",
     "Lichen" = "Mixed Lichen",
     "Attached" = "Litter"
    )
```

### create ID list

```{r}
view(mass_id6)

write.csv(mass_id6, "~/LTER-pluck/2024_pluck/processed_data/ag_pluck_id.csv", row.names = FALSE)
```

### combine

```{r, message=FALSE}
head(mass_id6)
#1134 of 7
head(mass_g3)
#4536 of 9

# join
mass3 <- left_join(mass_g3, mass_id6)
# Joining with by = join_by(block, treatment, surface_area_cm2, date_collected, species, tissue)
#4536 of 10
head(mass3)
```

### create functional group column

```{r}
unique(mass3$species)

# remove "loose litter", not counted consistently
mass4 <- mass3[!(mass3$species %in% "Loose"),]
#4464 of 10

# create Growth_form column
mass5 <- mass4 %>% 
  mutate(growth_form = case_when(
    startsWith(species, "Eriophorum vaginatum") ~ "Graminoid",
    startsWith(species, "Other Graminoid") ~ "Graminoid",
    startsWith(species, "Carex bigelowii") ~ "Graminoid",
    startsWith(species, "Betula nana") ~ "Deciduous",
    startsWith(species, "Salix pulchra") ~ "Deciduous",
    startsWith(species, "Vaccinium uliginosum") ~ "Deciduous",
    startsWith(species, "Other Deciduous") ~ "Deciduous",
    startsWith(species, "Ledum palustre") ~ "Evergreen",
    startsWith(species, "Vaccinium vitis-idaea") ~ "Evergreen",
    startsWith(species, "Cassiope tetragona") ~ "Evergreen",
    startsWith(species, "Other Evergreen") ~ "Evergreen",
    startsWith(species, "Rubus chamaemorus") ~ "Forb",
    startsWith(species, "Petasites frigidus") ~ "Forb",
    startsWith(species, "Other Forb") ~ "Forb",
    startsWith(species, "Sphagnum") ~ "Moss",
    startsWith(species, "Other Mosses") ~ "Moss",
    startsWith(species, "Mixed Lichen") ~ "Lichen",
    startsWith(species, "Litter") ~ "Litter"
    ))
#of 11
```

### add notes

```{r, message=FALSE, warning=FALSE}
# add notes
head(mass5)
mass5 <- mass5 %>%
  mutate(
    notes = case_when(
      (plot_tissue_id == "1736" & quad == "3") ~ "Suspected misidentified Eri vag rhizomes",
      (plot_tissue_id == "1739" & quad == "3") ~ "Suspected misidentified Litter"
    ))
#of 11
```

### final tidy

```{r}
col_order <- c("date_collected", "site", "block", "treatment", "quad", "surface_area_cm2", "species", "tissue", "growth_form", "plot_tissue_id", "mass_g", "notes")
mass5 <- mass5[, col_order]
```

### write final_data/ag_mass.csv

```{r}
view(mass5)
#4464 of 12

write.csv(mass5, "~/LTER-pluck/2024_pluck/processed_data/final_data/ag_mass.csv", row.names=FALSE)
```


## Belowground

### read in data

```{r, message=FALSE, warning=FALSE}
bg <- read_excel("~/LTER-pluck/2024_pluck/raw_data/Root_data_collection_mass.xlsx",
                 col_types = c("numeric", "text", "numeric", "text", "numeric", "numeric", "date", "date", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "text"))
#146 obs, 15 variables
#lots of columns that were numbers were being read as characters, appearing in scientific notation, etc
head(bg)
```

### tidy columns and rows

```{r}
bg <- clean_names(bg)

# remove unwanted columns
colnames(bg)

bg1 <- select(bg, -c("date_sorted"))
#146 obs, 14 variables

# remove 2 empty rows
bg2 <- bg1[-c(145:146),]
#144

# add site column
bg2$site <- c("MAT06")
#of 137
```

### separate masses and IDs

```{r, message=FALSE, warning=FALSE}
bg_names <- select(bg2, -ends_with("mass_g"), -c("date_collected"))
#144, 11
head(bg_names)

bg_mass <- select(bg2, -ends_with("bag_number"))
#144, 12
head(bg_mass)
```

### create longform mass table for publishing

```{r}
# wide2long
colnames(bg_mass)

bg_mass1 <- bg_mass %>% pivot_longer(cols = ends_with("_mass_g"), names_to = "root_type", values_to = "mass_g")
#432 of 10

bg_mass1$root_type <- 
  sub("_root_mass_g","",bg_mass1$root_type)
```

### sort IDs into one column

```{r}
colnames(bg_names)

bg_names1 <- select(bg_names, -c("core_surface_area_cm2","depth_cm","root_notes"))
#144 of 8

#wide2long
head(bg_names1)
bg_names2 <- bg_names1 %>% pivot_longer(cols = ends_with("_bag_number"), names_to = "root_type", values_to = "plot_root_id")
#432 of 7

bg_names3 <- bg_names2 %>% pivot_wider(names_from = quad, values_from = plot_root_id)
#108 of 9

# create ID for plot, ignore quads
bg_names3$plot_root_id <- coalesce(bg_names3$`1`, bg_names3$`2`, bg_names3$`3`, bg_names3$`4`)
#108 of 10
# ID is the first appearance of a bag number

bg_names3$root_type <- 
  sub("_root_bag_number","",bg_names3$root_type)

# remove quads
bg_names4 <- bg_names3 %>% 
  select(-c('1', '2', '3', '4'))
#of 6

head(bg_names4)
```

### create ID list

```{r}
view(bg_names4)
write.csv(bg_names4, "~/LTER-pluck/2024_pluck/processed_data/bg_pluck_id.csv", row.names = FALSE)
```

### combine

```{r, message=FALSE, warning=FALSE}
head(bg_names4)
#108 of 7
head(bg_mass1)
#432 of 11

bg3 <- left_join(bg_mass1, bg_names4)
# join_by(block, treatment, horizon, date_collected, site, root_type)
#432 of 12
```

### corrections

```{r}
# tidying notes
unique(bg3$root_notes)

# "First with new corer" 
# "Bag number is not mistake"
# "Sorting date"
# all organizational for sorting, not relevant to data

bg4 <- transform(bg3, root_notes = replace(root_notes, startsWith(root_notes, "First with new corer"), NA))
bg4 <- transform(bg4, root_notes = replace(root_notes, startsWith(root_notes, "Bag number"), NA))
bg4 <- transform(bg4, root_notes = replace(root_notes, startsWith(root_notes, "Sorting date"), NA))
```

### final tidy

```{r}
# reorder columns
colnames(bg4)
col_order <- c("date_collected", "site", "block", "treatment", "quad", "horizon", "core_surface_area_cm2", "depth_cm", "root_type", "plot_root_id", "mass_g", "root_notes")
bg4 <- bg4[, col_order]
```

### write final_data/bg_mass.csv

```{r}
view(bg4)
#432 of 12

write.csv(bg4, "~/LTER-pluck/2024_pluck/processed_data/final_data/bg_mass.csv", row.names=FALSE)
```


## Total Cores

### read in data

```{r, message=FALSE, warning=FALSE}
# read in files from /raw_data/ folder
tc <- read_excel("~/LTER-pluck/2024_pluck/raw_data/soil_data_total_cores_jm.xlsx", col_types = c("numeric", "text", "numeric", "text", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "date", "date", "numeric", "numeric", "numeric", "text"))
#83 obs, 16 variables
head(tc)
```

### tidy columns

```{r}
tc <- clean_names(tc)

colnames(tc)
# not really relevant other than for our organization during pluck
tc1 <- select(tc, -c("date_sorted", "sample_type"))
#83 obs, 14 variables

# add site column
tc1$site <- c("MAT06")
#of 15
```

### corrections

```{r}
# duplicate corrections
# B3 F5 Q4 two live, org and min, unclear which was used, kept A because it has depths and B mineral contained org
# B3 F0.5 Q4_defunct, mislabeled
# B2 F10 Q3 org labeled min, no record of 3B, just used 3A
# B2 F5 Q3 used redo 3B, excluded original 3A
# B2 F1 Q3 used redo 3B, excluded original 3A
# B2 F0.5 Q2 used redo 2B, excluded original 2A
# B2 CT Q4 no record of 4B, just used 4A
# B1 F10 Q1 no record of 1B, just used 1A
# B1 F2 Q3 no record of 3B, just used 3A
# B1 F1 Q3 no notes about two org samples existing, removed second with no depth
# B1 F0.5 Q4_defunct, mislabeled
tc2 <- tc1[-c(79, 65, 56, 50, 41, 35, 33, 25, 18, 13, 8),]
#72 obs, dropped 15

# tidying notes
unique(tc2$soil_notes)
# the info from "mineral is", "depth to permafrost", and "total depth to perm" got its own column, so removed from notes
# sludge cores were removed, so added explanation to redo

tc3 <- transform(tc2, soil_notes = replace(soil_notes, startsWith(soil_notes, "Mineral is"), NA))
tc3 <- transform(tc3, soil_notes = replace(soil_notes, startsWith(soil_notes, "Depth to permafrost"), NA))
tc3 <- transform(tc3, soil_notes = replace(soil_notes, startsWith(soil_notes, "Total depth"), NA))
tc3 <- transform(tc3, soil_notes = replace(soil_notes, startsWith(soil_notes, "This was the redo of 3A"), "This was a redo of 3A, which had a lot of sludge"))
tc3 <- transform(tc3, soil_notes = replace(soil_notes, startsWith(soil_notes, "This was a redo of 3A"), "This was a redo of 3A, which had a lot of sludge"))
tc3 <- transform(tc3, soil_notes = replace(soil_notes, startsWith(soil_notes, "This was the redo of 2A"), "This was a redo of 2A, which had a lot of sludge"))
```

### final tidy

```{r}
# reorder columns
colnames(tc3)
col_order <- c("date_collected", "site", "block", "treatment", "quad", "core_surface_area_cm2", "organic_core_depth_cm", "mineral_core_depth_cm", "depth_from_mineral_to_permafrost_cm", "depth_of_core_to_permafrost_cm", "ice_depth_accidentally_cored_cm", "dry_mass_aboveground_g", "dry_mass_organic_horizon_g", "dry_mass_mineral_horizon_g", "soil_notes")
tc3 <- tc3[, col_order]
```

### write final_data/tc_mass_quad.csv

```{r}
view(tc3)
#72 of 15

write.csv(tc3, "~/LTER-pluck/2024_pluck/processed_data/final_data/tc_mass.csv", row.names = FALSE)
```
