---
title: "Pluck cleaning"
format: html
author: Savannah Kjaer
date: 03-09-2026
---

## Set up

### load packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(readxl)
library(here)
```

## Aboveground

### read in data

```{r, message=FALSE, warning=FALSE}
# read in files from /raw_data/ folder
mass <- read_excel("~/LTER-pluck/raw_data/Shoot_data_collection_mass.xlsx")
#74 obs, 141 variables
head(mass)
```

### tidy columns and rows

```{r, message=FALSE}
# drop columns we don't want, organizational not informational
mass1 <- select(mass, 
                -c("Date_sorted","Plucker","Balance","Weigher","Notes"))
#74 obs, 136 variables

# drop empty rows at the bottom
mass2 <- mass1[-c(73:74),]
#72 obs of 136 variables
```

### separate mass and IDs

```{r}
# remove mass columns
mass_id <- select(mass2, -ends_with("mass_g"))
#72 obs, 73 variables

# remove other species column
mass_id1 <- select(mass_id, -ends_with("Species"))
#72 obs, 68 variables

# remove name columns
mass_g <- select(mass2, -ends_with("bag_number"))
#72 obs, 73 variables
mass_g1 <- select(mass_g, -ends_with("Species"))
#68 variables
```

### classes to numeric

```{r, warning=FALSE, output=FALSE}
# some mass columns as characters not numbers
# turns the "missing" values to NA's
cols.num <- c(6:68)
mass_g1[cols.num] <- sapply(mass_g1[cols.num],as.numeric)

# check
sapply(mass_g1, class)
```

### create longform mass table for publishing

```{r}
# wide2long
mass_g2 <- mass_g1 %>%
  pivot_longer(cols = c(6:68),
               names_to = "Tissue",
               values_to = "Mass_g")
#4536 obs, 7 variables

# remove "_mass_g" from tissue col
mass_g2$Tissue <- 
  sub("_mass_g","",mass_g2$Tissue)

# separate values in Tissue column into Species and Tissue
mass_g3 <- mass_g2 %>%
  separate(Tissue, c("Species", "Tissue"))
#4536 of 8
```

### sort IDs into one per plot

```{r}
# wide2long
mass_id2 <- mass_id1 %>% 
  pivot_longer(cols = c(ends_with('_bag_number')),
               names_to = "Tissue",
               values_to = "Plot_tissue_ID")
#4536 obs, 7 variables

# put all quads into one row per plot
mass_id3<-mass_id2 %>%
  pivot_wider(names_from = Quad,
              values_from = Plot_tissue_ID)
#1134 of 9

# create one ID per plot, ignore quads

# "NA" wasn't NA
mass_id4 <- mass_id3 %>%
  mutate(across(c(`1`, `2`, `3`, `4`), ~na_if(., "NA")))

# ID is the first appearance of a bag number
# if Q1 exists, uses that number, then tries Q2, Q3, Q4
mass_id4$Plot_tissue_ID <- coalesce(mass_id4$`1`, mass_id4$`2`, mass_id4$`3`, mass_id4$`4`)

# remove "_bag_number" in tissue column
mass_id4$Tissue <- 
  sub("_bag_number","",mass_id4$Tissue) 

# separate values in Tissue column into Species and Tissue
mass_id5 <- mass_id4 %>%
  separate(Tissue, c("Species", "Tissue"))
#1134 of 11

# remove quads
mass_id6 <- mass_id5[,-c(7:10)]
#1134 of 7

# now just one ID number for a plot
head(mass_id6)
```

### corrections

```{r}
# these ID's only correspond with B1 CT Q3&4, which were the burnt samples that will be excluded
# make them NA's for Q1&2
mass_id6$Plot_tissue_ID[mass_id6$Plot_tissue_ID == "204"] <- NA
mass_id6$Plot_tissue_ID[mass_id6$Plot_tissue_ID == "205"] <- NA
mass_id6$Plot_tissue_ID[mass_id6$Plot_tissue_ID == "258"] <- NA
mass_id6$Plot_tissue_ID[mass_id6$Plot_tissue_ID == "59"] <- NA

# two IDs from AG mass data so one renamed to another quad ID
mass_id6$Plot_tissue_ID[mass_id6$Plot_tissue_ID == "1152" & mass_id6$Treatment == "F5"] <- "1601"
```

### create ID list

```{r}
view(mass_id6)
write.csv(mass_id6, "~/LTER-pluck/processed_data/ag_pluck_id.csv", row.names = FALSE)
```

### combine

```{r, message=FALSE}
head(mass_id6)
#1134 of 7
head(mass_g3)
#4536 of 8

# join
mass3 <- left_join(mass_g3, mass_id6)
# Joining with by = join_by(Block, Treatment, Surface_area_cm2, Date_collected, Species, Tissue)
#4536 of 9
head(mass3)
```

### create functional group column

```{r}
unique(mass3$Species)

# remove "loose litter", not counted consistently
mass4 <- mass3[!(mass3$Species %in% "Loose"),]
#4464 of 9

#create Growth_form column
mass5 <- mass4 %>% mutate(Growth_form = case_when(
    startsWith(Species, "Erivag") ~ "Graminoid",
    startsWith(Species, "OtherGraminoid") ~ "Graminoid",
    startsWith(Species, "Carbig") ~ "Graminoid",
    startsWith(Species, "Betnan") ~ "Deciduous",
    startsWith(Species, "Salpul") ~ "Deciduous",
    startsWith(Species, "Vaculi") ~ "Deciduous",
    startsWith(Species, "OtherDeciduous") ~ "Deciduous",
    startsWith(Species, "Ledpal") ~ "Evergreen",
    startsWith(Species, "Vacvit") ~ "Evergreen",
    startsWith(Species, "Castet") ~ "Evergreen",
    startsWith(Species, "OtherEvergreen") ~ "Evergreen",
    startsWith(Species, "Rubcha") ~ "Forb",
    startsWith(Species, "Petfri") ~ "Forb",
    startsWith(Species, "OtherForb") ~ "Forb",
    startsWith(Species, "Sphagnum") ~ "Moss",
    startsWith(Species, "OtherMoss") ~ "Moss",
    startsWith(Species, "Lichen") ~ "Lichen",
    startsWith(Species, "Attached") ~ "Litter"
    ))
#of 10
```

### add notes

```{r, message=FALSE, warning=FALSE}
# add notes
head(mass5)
mass5 <- mass5 %>%
  mutate(
    Notes = case_when(
      (Plot_tissue_ID == "1736" & Quad == "3") ~ "Suspected misidentified Eri vag rhizomes",
      (Plot_tissue_ID == "1739" & Quad == "3") ~ "Suspected misidentified Litter"
    ))
#of 11
```

### write final_data/ag_mass.csv

```{r}
view(mass5)
#4464 of 11

write.csv(mass5, "~/LTER-pluck/processed_data/final_data/ag_mass.csv", row.names=FALSE)
```


## Belowground

### read in data

```{r, message=FALSE, warning=FALSE}
bg <- read_excel("~/LTER-pluck/raw_data/Root_data_collection_mass.xlsx",
                 col_types = c("numeric", "text", "numeric", "text", "numeric", "numeric", "date", "date", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "text"))
#146 obs, 15 variables
#lots of columns that were numbers were being read as characters, appearing in scientific notation, etc
head(bg)
```

### tidy columns and rows

```{r}
# remove unwanted columns
colnames(bg)

bg1 <- select(bg, -c("Date_sorted"))
#146 obs, 14 variables

# remove 2 empty rows
bg2 <- bg1[-c(145:146),]
#144
```

### separate masses and IDs

```{r, message=FALSE, warning=FALSE}
bg_names <- select(bg2, -ends_with("mass_g"))
#144, 11
head(bg_names)

bg_mass <- select(bg2, -ends_with("bag_number"))
#144, 11
head(bg_mass)
```

### create longform mass table for publishing

```{r}
# wide2long
colnames(bg_mass)

bg_mass1 <- bg_mass %>% pivot_longer(cols = c(8:10), names_to = "Root_type", values_to = "Mass_g")
#432 of 10

bg_mass1$Root_type <- 
  sub("_root_mass_g","",bg_mass1$Root_type)
```

### sort IDs into one column

```{r}
colnames(bg_names)

bg_names1 <- select(bg_names, -c("Core_surface_area_cm2","Depth_cm","Root_notes"))
#144 of 8

#wide2long
head(bg_names1)
bg_names2 <- bg_names1 %>% pivot_longer(cols = c(6:8), names_to = "Root_type", values_to = "Plot_root_ID")
#432 of 7

bg_names3<-bg_names2 %>% pivot_wider(names_from = Quad, values_from = Plot_root_ID)
#108 of 9

# create ID for plot, ignore quads
bg_names3$Plot_root_ID <- coalesce(bg_names3$`1`, bg_names3$`2`, bg_names3$`3`, bg_names3$`4`)
#108 of 10
# ID is the first appearance of a bag number

bg_names3$Root_type <- 
  sub("_root_bag_number","",bg_names3$Root_type)

# remove quads
bg_names4 <- bg_names3[,-c(6:9)]
head(bg_names4)
```

### create ID list

```{r}
view(bg_names4)
write.csv(bg_names4, "~/LTER-pluck/processed_data/bg_pluck_id.csv", row.names = FALSE)
```

### combine

```{r, message=FALSE, warning=FALSE}
head(bg_names4)
#108 of 6
head(bg_mass1)
#432 of 10

bg3<-left_join(bg_mass1, bg_names4)
# join_by(Block, Treatment, Horizon, Date_collected, Root_type)
#432 of 11

# reorder columns
colnames(bg3)
col_order <- c("Block", "Treatment", "Quad", "Horizon", "Core_surface_area_cm2", "Depth_cm", "Date_collected", "Root_type", "Mass_g", "Plot_root_ID", "Root_notes")
bg3 <- bg3[, col_order]

```

### corrections

```{r}
# tidying notes
unique(bg3$Root_notes)

# "First with new corer" 
# "Bag number is not mistake"
# "Sorting date"
# all organizational for sorting, not relevant to data

bg4 <- transform(bg3, Root_notes = replace(Root_notes, startsWith(Root_notes, "First with new corer"), NA))
bg4 <- transform(bg4, Root_notes = replace(Root_notes, startsWith(Root_notes, "Bag number"), NA))
bg4 <- transform(bg4, Root_notes = replace(Root_notes, startsWith(Root_notes, "Sorting date"), NA))
```

### write final_data/bg_mass.csv

```{r}
view(bg4)
#432 of 11

write.csv(bg4, "~/LTER-pluck/processed_data/final_data/bg_mass.csv", row.names=FALSE)
```


## Total Cores

### read in data

```{r, message=FALSE, warning=FALSE}
# read in files from /raw_data/ folder
tc <- read_excel("~/LTER-pluck/raw_data/Soil_data_total_cores_jm.xlsx")
#83 obs, 16 variables
head(tc)
```

### tidy columns

```{r}
colnames(tc)
# not really relevant other than for our organization during pluck
tc1 <- select(tc, -c("Date_sorted", "Sample_Type", "Date_sorted"))
#83 obs, 14 variables
```

### corrections

```{r}
# duplicate corrections
# B3 F5 Q4 two live, org and min, unclear which was used, kept A because it has depths and B mineral contained org
# B3 F0.5 Q4_defunct, mislabeled
# B2 F10 Q3 org labeled min, no record of 3B, just used 3A
# B2 F5 Q3 used redo 3B, excluded original 3A
# B2 F1 Q3 used redo 3B, excluded original 3A
# B2 F0.5 Q2 used redo 2B, excluded original 2A
# B2 CT Q4 no record of 4B, just used 4A
# B1 F10 Q1 no record of 1B, just used 1A
# B1 F2 Q3 no record of 3B, just used 3A
# B1 F1 Q3 no notes about two org samples existing, removed second with no depth
# B1 F0.5 Q4_defunct, mislabeled
tc2 <- tc1[-c(79, 65, 56, 50, 41, 35, 33, 25, 18, 13, 8),]
#72 obs, dropped 15

# tidying notes
unique(tc2$Soil_notes)
# the info from "mineral is", "depth to permafrost", and "total depth to perm" got its own column, so removed from notes
# sludge cores were removed, so added explanation to redo

tc3 <- transform(tc2, Soil_notes = replace(Soil_notes, startsWith(Soil_notes, "Mineral is"), NA))
tc3 <- transform(tc3, Soil_notes = replace(Soil_notes, startsWith(Soil_notes, "Depth to permafrost"), NA))
tc3 <- transform(tc3, Soil_notes = replace(Soil_notes, startsWith(Soil_notes, "Total depth"), NA))
tc3 <- transform(tc3, Soil_notes = replace(Soil_notes, startsWith(Soil_notes, "This was the redo of 3A"), "This was a redo of 3A, which had a lot of sludge"))
tc3 <- transform(tc3, Soil_notes = replace(Soil_notes, startsWith(Soil_notes, "This was a redo of 3A"), "This was a redo of 3A, which had a lot of sludge"))
tc3 <- transform(tc3, Soil_notes = replace(Soil_notes, startsWith(Soil_notes, "This was the redo of 2A"), "This was a redo of 2A, which had a lot of sludge"))
```

### write final_data/tc_mass_quad.csv

```{r}
view(tc3)
#72 of 14

write.csv(tc3, "~/LTER-pluck/processed_data/final_data/tc_mass.csv", row.names = FALSE)
```

