---
title: "EA_cleaning"
format: html
---

## Set up

### load packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(readxl)
library(here)
```

## Aboveground

### read in data

```{r, message=FALSE}
cn <- read_excel("~/LTER-pluck/raw_data/CN_results2.xlsx")
view(cn)
#953 of 17 variables

id <- read_csv("~/LTER-pluck/processed_data/ag_pluck_id.csv")
#1134 of 7
```

### tidy columns and rows

```{r, message=FALSE}
colnames(cn)
# removing unnecessary columns
cn1 <- select(cn, -c("Group", "Sample name", "Inj Date", "Inj Time", "Type", "Weight (mg)", "Protein Factor", "Humidity %", "AS", "Vial"))
#953 of 7

cn1$Carbon <- as.numeric(cn1$Carbon)
cn1$Nitrogen <- as.numeric(cn1$Nitrogen)
cn1$Filename <- as.character(cn1$Filename)
names(cn1)[names(cn1) == 'Sample type'] <- "Sample_type"
names(cn1)[names(cn1) == "ID"] <- "Plot_tissue_ID"
```

### separate AG masses from others

```{r}
cn_ag <- filter(cn1, Sample_type == "AG")
view(cn_ag)
#707
```

### corrections

```{r}
# two IDs from AG mass data so one renamed to another quad ID
cn_ag$Plot_tissue_ID[cn_ag$Filename == "09042025013"] <- "1601"

# looking into IDs that appeared in EA results that are not ID names
# 602, 685, 1548, 1549, 1084
# 602 was mistyped, paper datasheet say 662
# 685 was mistyped, paper datasheet say 658
# 1548 was mistyped, paper datasheet say 1506
# 1549 was mistyped, paper datasheet say 1507
# 223 was mistyped, paper datasheet say 222
# 1084 unclear what sample it was

cn_ag$Plot_tissue_ID[cn_ag$Filename == "04042025044"] <- "662"
cn_ag$Plot_tissue_ID[cn_ag$Filename == "04042025042"] <- "658"
cn_ag$Plot_tissue_ID[cn_ag$Filename == "09042025017"] <- "1506"
cn_ag$Plot_tissue_ID[cn_ag$Filename == "09042025018"] <- "1507"
cn_ag$Plot_tissue_ID[cn_ag$Filename == "03042025105"] <- "222"
cn_ag$Plot_tissue_ID[cn_ag$Filename == "24032025073"] <- "NA"
```

### combine

```{r, message=FALSE}
head(cn_ag)
#707 obs of 7

head(id)
#1134 of 7 
id$Plot_tissue_ID <- as.character(id$Plot_tissue_ID)

cn2 <- left_join(cn_ag, id)
# Joining with `by = join_by(Plot_tissue_ID)`

view(cn2)
#707 of 13
```

### error check

```{r, message=FALSE}
cn2.na <- subset(cn2, (is.na(cn2$Block)))
#107 obs

# remove check results
cn3 <- cn2 %>%
  filter(!grepl("Check", Plot_tissue_ID))
#649

# "NA" wasn't NA
cn3$Plot_tissue_ID[cn3$Plot_tissue_ID == "NA"] <- NA

cn3.na <- subset(cn3, (is.na(cn3$Plot_tissue_ID)))
#14
# checked, can all be removed 

cn4 <- cn3 %>%
  filter(!(Plot_tissue_ID %in% c(NA)))
#635, good!!

cn4.na <- subset(cn4, (is.na(cn4$Block)))
```

### burnt samples

Looked at values of burnt against unburnt samples, not terribly different but I think it's best to remove them entirely, rather than keep even potentially flawed samples. This removes all samples from B1 F0.5, and B1 CT Q3&4.

```{r}
# checking burnt B1 against unburnt B2&3
test <- cn4 %>%
  filter(Treatment %in% c("F0.5"))

test %>%
  ggplot() +
  geom_boxplot(aes(x=Tissue, y=Carbon, col=Notes)) + 
  facet_wrap(~Species)
test %>%
  ggplot() +
  geom_boxplot(aes(x=Tissue, y=Nitrogen, col=Notes)) + 
  facet_wrap(~Species)
```

### remove burnt samples

```{r}
burnt <- cn4 %>%
  filter(Notes %in% c("burnt sample"))
# all burnt, should be removed

cn5 <- cn4 %>%
  filter(!(Notes %in% c("burnt sample")))

view(cn5)
#562 obs
```

### create growth form column

```{r}
#create Growth_form column
cn6 <- cn5 %>% mutate(Growth_form = case_when(
    startsWith(Species, "Erivag") ~ "Graminoid",
    startsWith(Species, "OtherGraminoid") ~ "Graminoid",
    startsWith(Species, "Carbig") ~ "Graminoid",
    startsWith(Species, "Betnan") ~ "Deciduous",
    startsWith(Species, "Salpul") ~ "Deciduous",
    startsWith(Species, "Vaculi") ~ "Deciduous",
    startsWith(Species, "OtherDeciduous") ~ "Deciduous",
    startsWith(Species, "Ledpal") ~ "Evergreen",
    startsWith(Species, "Vacvit") ~ "Evergreen",
    startsWith(Species, "Castet") ~ "Evergreen",
    startsWith(Species, "OtherEvergreen") ~ "Evergreen",
    startsWith(Species, "Rubcha") ~ "Forb",
    startsWith(Species, "Petfri") ~ "Forb",
    startsWith(Species, "OtherForb") ~ "Forb",
    startsWith(Species, "Sphagnum") ~ "Moss",
    startsWith(Species, "OtherMoss") ~ "Moss",
    startsWith(Species, "Lichen") ~ "Lichen",
    startsWith(Species, "Attached") ~ "Litter"
    ))
#of 14
```

### final tidy

```{r}
head(cn6)

cn7 <- select(cn6, -c("Filename", "Date", "Sample_type"))
col_order <- c("Block", "Treatment", "Surface_area_cm2", "Date_collected", "Species", "Tissue", "Plot_tissue_ID", "Nitrogen", "Carbon", "Growth_form", "Notes")
cn7 <- cn7[, col_order]

names(cn7)[names(cn7) == 'Nitrogen'] <- "Nitrogen_%"
names(cn7)[names(cn7) == 'Carbon'] <- "Carbon_%"

cn7$Surface_area_cm2 <- c(1600)
# since it's 4 quads combined, but maybe this column is better just removed?
```

### write final_data/ag_cn.csv

```{r}
view(cn7)

write.csv(cn7, "~/LTER-pluck/processed_data/final_data/ag_cn.csv", row.names=FALSE)
```

## Belowground

### read in data

```{r, message=FALSE}
id_bg <- read_csv("~/LTER-pluck/processed_data/bg_pluck_id.csv")
#108 of 6
```

### separate BG masses from others

```{r}
names(cn1)[names(cn1) == "Plot_tissue_ID"] <- "Plot_root_ID"

cn_bg <- filter(cn1, Sample_type == "BG")
view(cn_bg)
#80
```

### corrections

```{r}
# looking into IDs that appeared in EA results that are not ID names
# 395 unclear what sample it was

cn_bg$Plot_root_ID[cn_bg$Filename == "11042025025"] <- "NA"
cn_bg$Plot_root_ID[cn_bg$Filename == "24032025041"] <- "175"
```

### combine

```{r, message=FALSE}
head(id_bg)
#108 of 6
id_bg$Plot_root_ID <- as.character(id_bg$Plot_root_ID)

cn_bg2 <- left_join(cn_bg, id_bg)
# Joining with `by = join_by(Plot_root_ID)`

view(cn_bg2)
#80 of 12
```

### error check

```{r, message=FALSE}
cn2.na <- subset(cn_bg2, (is.na(cn_bg2$Block)))
#7 obs

# remove check results
cn_bg3 <- cn_bg2 %>%
  filter(!grepl("Check", Plot_root_ID))
#74

# "NA" wasn't NA
cn_bg3$Plot_root_ID[cn_bg3$Plot_root_ID == "NA"] <- NA

cn_bg3.na <- subset(cn_bg3, (is.na(cn_bg3$Plot_root_ID)))
#1
# checked, can be removed 

cn_bg4 <- cn_bg3 %>%
  filter(!(Plot_root_ID %in% c(NA)))
#73, good!!

cn_bg4.na <- subset(cn_bg4, (is.na(cn_bg4$Block)))
```

### final tidy

```{r}
head(cn_bg4)

cn_bg5 <- select(cn_bg4, -c("Filename", "Date", "Sample_type"))
col_order <- c("Block", "Treatment", "Date_collected", "Horizon", "Root_type", "Plot_root_ID", "Nitrogen", "Carbon", "Notes")
cn_bg5 <- cn_bg5[, col_order]

names(cn_bg5)[names(cn_bg5) == 'Nitrogen'] <- "Nitrogen_%"
names(cn_bg5)[names(cn_bg5) == 'Carbon'] <- "Carbon_%"
```

### write final_data/ag_cn.csv

```{r}
view(cn_bg5)

write.csv(cn_bg5, "~/LTER-pluck/processed_data/final_data/bg_cn.csv", row.names=FALSE)
```

## Total Cores

### separate BG masses from others

```{r}
names(cn1)[names(cn1) == "Plot_root_ID"] <- "Plot_horizon_ID"

cn_tc <- filter(cn1, Sample_type == "Total Cores")
view(cn_tc)
#80
```

### corrections

```{r}
cn_tc$Plot_horizon_ID <- gsub("-T", "-M", cn_tc$Plot_horizon_ID)
# not sure why mineral was labeled as T?
```

### combine

```{r, message=FALSE}
head(cn_tc)
#44 of 7

cn_bg2 <- left_join(cn_bg, id_bg)
# Joining with `by = join_by(Plot_root_ID)`

# START HERE
# add ID to mass data
# add block, trt, what else to cn results

view(cn_bg2)
#80 of 12
```

### error check

```{r, message=FALSE}
cn2.na <- subset(cn_bg2, (is.na(cn_bg2$Block)))
#7 obs

# remove check results
cn_bg3 <- cn_bg2 %>%
  filter(!grepl("Check", Plot_root_ID))
#74

# "NA" wasn't NA
cn_bg3$Plot_root_ID[cn_bg3$Plot_root_ID == "NA"] <- NA

cn_bg3.na <- subset(cn_bg3, (is.na(cn_bg3$Plot_root_ID)))
#1
# checked, can be removed 

cn_bg4 <- cn_bg3 %>%
  filter(!(Plot_root_ID %in% c(NA)))
#73, good!!

cn_bg4.na <- subset(cn_bg4, (is.na(cn_bg4$Block)))
```

### final tidy

```{r}
head(cn_bg4)

cn_bg5 <- select(cn_bg4, -c("Filename", "Date", "Sample_type"))
col_order <- c("Block", "Treatment", "Date_collected", "Horizon", "Root_type", "Plot_root_ID", "Nitrogen", "Carbon", "Notes")
cn_bg5 <- cn_bg5[, col_order]

names(cn_bg5)[names(cn_bg5) == 'Nitrogen'] <- "Nitrogen_%"
names(cn_bg5)[names(cn_bg5) == 'Carbon'] <- "Carbon_%"
```

### write final_data/ag_cn.csv

```{r}
view(cn_bg5)

write.csv(cn_bg5, "~/LTER-pluck/processed_data/final_data/bg_cn.csv", row.names=FALSE)
```



