---
title: "Pluck cleaning"
format: html
author: Savannah Kjaer
date: 03-09-2026
---

# aboveground

## Set up

### load packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(readxl)
library(here)
```

### read in data

```{r, message=FALSE, warning=FALSE}
# read in files from /raw_data/ folder
mass <- read_excel("~/LTER-pluck/raw_data/Shoot_data_collection_mass.xlsx")
#74 obs, 141 variables
head(mass)
```

## Tidy

### tidy columns and rows

```{r, message=FALSE}
colnames(mass)
# drop columns we don't want
mass1 <- select(mass, 
                -c("Date_collected","Date_sorted","Plucker","Balance","Weigher","Notes"))
#74 obs, 135 variables

# drop empty rows at the bottom
mass2 <- mass1[-c(73:74),]
#72 obs of 135 variables
```

### separate mass and IDs

```{r}
# remove mass columns
mass_id <- select(mass2, -ends_with("mass_g"))
#72 obs, 72 variables

# remove other species column
mass_id1 <- select(mass_id, -ends_with("Species"))
#72 obs, 67 variables

### go back and add other species later 

# remove name columns
mass_g <- select(mass2, -ends_with("bag_number"))
#72 obs, 72 variables
mass_g1 <- select(mass_g, -ends_with("Species"))
#67 variables
head(mass_g1)
```

### sort IDs into one per plot

```{r}
# preserve proper copy, 72 of 67
mass_id_copy<- mass_id1

# wide2long
mass_id2 <- mass_id1 %>% 
  pivot_longer(cols = c(ends_with('_bag_number')),
               names_to = "Tissue",
               values_to = "ID")
#4536 obs, 6 variables

# put all quads into one row per plot
mass_id3<-mass_id2 %>%
  pivot_wider(names_from = Quad,
              values_from = ID)
#1134 of 9

# create one ID per plot, ignore quads
mass_id3[mass_id3 == "NA"] <- NA
# "NA" wasn't NA

mass_id3$ID <- coalesce(mass_id3$`1`, mass_id3$`2`, mass_id3$`3`, mass_id3$`4`)
# ID is the first appearance of a bag number
# if Q1 exists, uses that number, then tries Q2, Q3, Q4

#remove "_bag_number" in tissue column
mass_id3$Tissue <- 
  sub("_bag_number","",mass_id3$Tissue) 

# separate values in Tissue column into Species and Tissue
mass_id4 <- mass_id3 %>%
  separate(Tissue, c("Species", "Tissue"))
#1134 of 10

# remove Q1-4
mass_id5 <- mass_id4[,-c(6:9)]
#1134 of 6
# now just one ID number for a plot
head(mass_id5)

write.csv(mass_id5, "~/LTER-pluck/processed_data/ag_ids.csv", row.names=FALSE)
```

## COME BACK TO THIS

how to deal with burnt samples

```{r}
# these ID's only correspond with Q3&4, they are NA's for Q1&2
# because of the burnt 
#mass_id5$ID[mass_id5$ID == "204" & mass_id5$Block == "1"] <- NA
#mass_id5$ID[mass_id5$ID == "205" & mass_id5$Block == "1"] <- NA
#mass_id5$ID[mass_id5$ID == "258" & mass_id5$Block == "1"] <- NA
#mass_id5$ID[mass_id5$ID == "59" & mass_id5$Block == "1"] <- NA
```

### classes to numeric

```{r, warning=FALSE, output=FALSE}
# preserve proper copy, 72 of 67
mass_g_copy<-mass_g1

# some values characters not numbers
# turns the "missing" values to NA's
cols.num <- c(5:67)
mass_g1[cols.num] <- sapply(mass_g1[cols.num],as.numeric)
# check
sapply(mass_g1, class)
```

### create longform mass table for publishing

```{r}
# wide2long
mass_g2 <- mass_g1 %>%
  pivot_longer(cols = c(5:67),
               names_to = "Tissue",
               values_to = "Mass_g")
#4536 obs, 6 variables

# remove "_mass_g" from tissue col
mass_g2$Tissue <- 
  sub("_mass_g","",mass_g2$Tissue)

# separate values in Tissue column inot Species and Tissue
mass_g3 <- mass_g2 %>%
  separate(Tissue, c("Species", "Tissue"))
#4536 of 7
head(mass_g3)

write.csv(mass_g3, "~/LTER-pluck/processed_data/final_data/ag_mass_quad.csv", row.names=FALSE)
```

### sum masses across quads for one value per plot

```{r}
# put all quads into one row per plot
mass_g4<-mass_g3 %>%
  pivot_wider(names_from = Quad, values_from = Mass_g)
#1134 of 9

# sum quads to combine with EA
# should we combine them or leave it off?? bc then the same data will be in 2 places
mass_g_sum <- mass_g4 %>%
rowwise() %>%
mutate(Total_mass_g = sum(across(c(6:9)), na.rm = T))
#1134 of 10

# remove Q1-4
# keep only one obs per row
mass_g_sum1 <- mass_g_sum[,-c(6:9)]
#1134 of 6
# now just one total mass per plot
head(mass_g_sum1)
```

### combine

```{r}
head(mass_id5)
#1134 of 6
head(mass_g_sum1)
#1134 of 6

# join
mass3 <- left_join(mass_id5, mass_g_sum1)
# Joining with by = join_by(Block, Treatment, Surface_area_cm2, Species, Tissue)
#1134 of 3
head(mass3)
```

### create functional group column

```{r}
unique(mass3$Species)

#remove "loose litter"
mass4 <- mass3[!(mass3$Species %in% "Loose"),]
#1116 of 7

#create Growth_form column
mass5 <- mass4 %>% mutate(Growth_form = case_when(
    startsWith(Species, "Erivag") ~ "Graminoid",
    startsWith(Species, "OtherGraminoid") ~ "Graminoid",
    startsWith(Species, "Carbig") ~ "Graminoid",
    startsWith(Species, "Betnan") ~ "Deciduous",
    startsWith(Species, "Salpul") ~ "Deciduous",
    startsWith(Species, "Vaculi") ~ "Deciduous",
    startsWith(Species, "OtherDeciduous") ~ "Deciduous",
    startsWith(Species, "Ledpal") ~ "Evergreen",
    startsWith(Species, "Vacvit") ~ "Evergreen",
    startsWith(Species, "Castet") ~ "Evergreen",
    startsWith(Species, "OtherEvergreen") ~ "Evergreen",
    startsWith(Species, "Rubcha") ~ "Forb",
    startsWith(Species, "Petfri") ~ "Forb",
    startsWith(Species, "OtherForb") ~ "Forb",
    startsWith(Species, "Sphagnum") ~ "Moss",
    startsWith(Species, "OtherMoss") ~ "Moss",
    startsWith(Species, "Lichen") ~ "Lichen",
    startsWith(Species, "Attached") ~ "Litter"
    ))
#1116 of 8

#Sample ID number appeared twice, 1152 
#counted the one that ran with samples from the same plot
mass5$ID[mass5$ID == "1152" & mass5$Species == "Erivag"] <- NA
head(mass5)

write.csv(mass5, "~/LTER-pluck/processed_data/ag_id_mass.csv", row.names=FALSE)
```


## Belowground

### read in data

```{r, message=FALSE, warning=FALSE}
bg <- read_excel("~/LTER-pluck/raw_data/Root_data_collection_mass.xlsx",
                 col_types = c("numeric", "text", "numeric", "text", "numeric", "numeric", "date", "date", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "text"))
#146 obs, 15 variables
#lots of columns that were numbers were being read as characters, appearing in scientific notation, etc
head(bg)
colnames(bg)
```

### tidy columns and rows

```{r}
# remove unwanted columns
bg1 <- select(bg, -c("Date_collected","Date_sorted"))
#146 obs, 13 variables

# remove 2 empty rows
bg2 <- bg1[-c(145:146),]
#144
```

### separate masses and IDs

```{r}
bg_names <- select(bg2, -ends_with("mass_g"))
#144, 10
head(bg_names)

bg_mass <- select(bg2, -ends_with("bag_number"))
#144, 10
head(bg_mass)
```

### sort IDs into one column

```{r}
bg_names_copy<-bg_names
# preserve proper copy, 144 of 10

colnames(bg_names)
bg_names1 <- select(bg_names, -c("Core_surface_area_cm2","Depth_cm","Root_notes"))
#144 of 7

#wide2long
bg_names2 <- bg_names1 %>% pivot_longer(cols = c(5:7), names_to = "Root_type", values_to = "ID")
#432 of 6

bg_names3<-bg_names2 %>% pivot_wider(names_from = Quad, values_from = ID)
#108 of 8

# create ID for plot, ignore quads
bg_names3$ID <- coalesce(bg_names3$`1`, bg_names3$`2`, bg_names3$`3`, bg_names3$`4`)
#108 of 9
# ID is the first appearance of a bag number

bg_names3$Root_type <- 
  sub("_root_bag_number","",bg_names3$Root_type)

# remove quads
bg_names4 <- bg_names3[,-c(5:8)]
head(bg_names4)

write.csv(bg_names4, "~/LTER-pluck/processed_data/bg_ids.csv", row.names=FALSE)
```

### create longform mass table for publishing

```{r}
# wide2long
bg_mass_copy<-bg_mass
# preserve proper copy, 144 of 11
colnames(bg_mass)

# calculate core volume
bg_mass$Core_volume_cm3<-with(bg_mass, Core_surface_area_cm2 * Depth_cm)

bg_mass1 <- bg_mass %>% pivot_longer(cols = c(7:9), names_to = "Root_type", values_to = "Mass_g")
#432 of 10

bg_mass1$Root_type <- 
  sub("_root_mass_g","",bg_mass1$Root_type)
head(bg_mass1)

write.csv(bg_mass1, "~/LTER-pluck/processed_data/final_data/bg_mass_quad.csv", row.names=F)
```

### sum masses across quads

```{r}
# sum quads
bg_sum <- bg_mass1 %>% 
  group_by(Block, Treatment, Horizon, Root_type) %>%
  summarise( 
    Total_mass_g = sum(Mass_g, na.rm = TRUE), # summarizing quads for one per plot
    Core_surface_area_cm2 = mean(Core_surface_area_cm2, na.rm=TRUE), # should be the same within a plot
    # removed depth bc it varies with each q, mean or sum feels deceptive
    Total_core_volume_cm3 = sum(Core_volume_cm3, na.rm = TRUE)) #sum of all cores
view(bg_sum)
#108 of 7
```

### combine

```{r}
head(bg_names4)
#108 of 5
head(bg_sum)
#108 of 7

bg3<-left_join(bg_names4, bg_sum)
# join_by(Block, Treatment, Horizon, Root_type)
#108 obs of 8
head(bg3)

write.csv(bg3, "~/LTER-pluck/processed_data/bg_id_mass.csv", row.names=FALSE)
```


## Total Cores

### read in data

```{r, message=FALSE, warning=FALSE}
# read in files from /raw_data/ folder
tc <- read_excel("~/LTER-pluck/raw_data/Soil_data_total_cores_jm.xlsx")
#83 obs, 14 variables
head(tc)

```

